A Data-Driven Framework for Neural Field Modeling

### 2011-01-05 ltx/first_order_framework/LaTeX_Manuscript.tex
very minor edit

For models to be clinically viable, they must be patient-specific. A possible approach to achieve this would be to fit a general continuum neural field model, like the \citet{Wilson1973} (WC) or \citet{Amari1977} models, or a neural mass model like the \citet{Jansen1995} model, to patient-specific data. Fitting the neural models to individuals is a highly non-trivial task. Recently, however, this task has been approached from a number of standpoints. \dean{(Reviewer 1, additional comment d):) The first paper on patient specific modeling came from~\citet{Valdes1999}, where they fit a neural mass model of \citet{LopesDaSilva1976} and \citet{Zetterberg1978} \parham{using an extended Kalman filter--need to check (it's not extended Kalman filter I think-maybe just say using nonlinear Kalman filtering algorithm)}. This paper demonstrated for the first time the feasibility of assimilating data with neural models of EEG.}

@@

The support of the connectivity kernel can be inferred using spatial correlation analysis. The method presented in this paper is based on a similar published method, where the kernel support was estimated using correlation analysis for a linear system described by an IDE~\citep{Scerri2009}. To demonstrate the method for the neural field model we assume the sensors are infinitesimally close (continuous observation). The spatial cross-correlation function between consecutive observations is defined as 

@@

where $\mathcal{F}^{-1}$ denotes the inverse Fourier transform and $R_{y_{t},y_{t}}(\boldsymbol{\tau})$ is the auto-correlation of the observations, $\delta_K(\cdot)$ is the Kronecker delta function, and $\delta_D(\cdot)$ is the Dirac delta function.

We define the auto-correlation of the observations as

@@

This equation states that the support of the kernel can be estimated given knowledge of $\sigma_\varepsilon^2$. Furthermore, is we assume a local coupling (at $\boldsymbol\tau=0$) regime, either excitation or inhibition, than we don't even need to know $\sigma_\varepsilon^2$. The peak in the kernel at $\mathbf{r}=0$ will be bigger or smaller if we guess incorrectly and the kernel parameters can do the rest!

@@

The sigma vectors are propagated through the system equations and weighted to form the predicted mean and covariance. The weights are calculated by 
\begin{align}
	\mathbf W_0^{(m)}&=\frac{\lambda}{ n_x+\lambda} \\
	\mathbf W_0^{(c)}&=\frac{\lambda}{ n_x+\lambda}+(1-\alpha^2+\beta) \\
	\mathbf W_i^{(m)}&=\mathbf W_i^{(c)}=\frac{1}{2( n_x+\lambda)} \quad i=1, \dots, 2n_x, 
\end{align}
where the superscripts $m$ and $c$ denote mean and covariance and $\beta$ incorporates prior knowledge of the distribution of the states, $\mathbf{x}$ \dean{(for a Gaussian disturbance $\beta$ should be set to 2 \parham{\citep{Haykin2001}})}. Since the observation equation is linear (equation~\ref{ObservationEquation}), the standard Kalman Filter update equations are used to correct the predicted states. The state estimates from forward filtering are used to form a new set of sigma points for the smoother, as described above. The cross-covariance matrix of the states, $\mathbf M$, is also required for computing the smoother gain, as described in the summary of the steps in the URTSS algorithm is given in Table~\ref{tab:UKFAlgorithm}.

Although the system is nonlinear, the parameters of the system are linear with respect to the state. This is exploited by our procedure where the parameter estimation uses a least squares (LS) method that minimizes the sum of the squared errors (of a predicted state update) with each new estimate of the state vector sequence. To create the least squares parameter estimator, we first define the $n_x \times n_{\theta}$ matrix
\begin{equation}
	\mathbf{q}(\mathbf{x}_t) = \int_\Omega \boldsymbol{\Psi}(\mathbf{r}') f(\boldsymbol{\phi}^{\top}(\mathbf{r}')\mathbf{x}_t) \,\textrm{d}\mathbf{r}'.
\end{equation}
Now given a state estimate sequence from the initialization or an iteration of the URTSS, we can write
\begin{align*}
	 \hat{\mathbf x}^b_{1} &= \mathbf{q}(\hat{\mathbf x}^b_0) \boldsymbol{\theta}+\xi\hat{\mathbf x}^b_0+\mathbf e_0 \\
	\hat{\mathbf x}^b_{2} &= \mathbf{q}(\hat{\mathbf x}^b_1) \boldsymbol{\theta}+\xi\hat{\mathbf x}^b_1+\mathbf e_1  \\
	&\vdots& \\
	\hat{\mathbf x}^b_{T}&= \mathbf{q}(\hat{\mathbf x}^b_{T-1}) \boldsymbol{\theta}+\xi\hat{\mathbf x}^b_{T-1}+\mathbf e_{T-1}. 
\end{align*}
This can be written in the compact form
\begin{equation}
	\mathbf Z=\mathbf X \mathcal W+\mathbf{e}, 
\end{equation}
where
\begin{small}
\begin{equation*}
	\mathbf Z=\left[
	\begin{array}{cccc}
		\hat{\mathbf x}^b_{1}\\
		\hat{\mathbf x}^b_{2}\\
		\vdots\\
		\hat{\mathbf x}^b_{T}
	\end{array}
	\right],\quad \mathbf X=\left[
	\begin{array}{cccc}
		\mathbf q(\hat{\mathbf x}^b_0)& \hat{\mathbf x}^b_{0}\\
		\mathbf q(\hat{\mathbf x}^b_1)& \hat{\mathbf x}^b_{1}\\
		\vdots & \vdots\\
		\mathbf q(\hat{\mathbf x}^b_{T-1})& \hat{\mathbf x}^b_{T-1}
	\end{array}
	\right] 
\end{equation*}
\end{small}
and
\begin{small}
\begin{equation*}
\quad \mathcal W=\left[
	\begin{array}{cc}
		\boldsymbol{\theta} \\
		\xi
	\end{array}
	\right],\quad \mathbf{e}=\left[
	\begin{array}{cccc}
		\mathbf e_0\\
		\mathbf e_1\\
		\vdots\\
		\mathbf e_{T-1}
	\end{array}
	\right].
\end{equation*}
\end{small}
Following this, the LS parameter estimates \parham{\citep{Ljung1999}}, $ \hat{\mathcal{W}}$, are
\begin{equation}
	\hat{\mathcal{W}}=(\mathbf X^\top\mathbf X)^{-1}\mathbf X^\top\mathbf Z. 
\end{equation}

@@

\dean{The framework assumes that the parameters of the model (and brain) are stationary over the estimation period. This is a reasonable assumption if applying the framework to an evoked-potential experiment over a short time period in a controlled environment. However, it is well known that factors that influence excitability are non-stationary over longer time periods (see \citet{Ullah2010} for a nice example). Therefore, the estimation framework must be altered if it is to be applied to data over extended periods of time. A possible way to achieve this is to use a more sophisticated model that accounts for longer term time dependancies (higher temporal order). Alternatively, the parameters of the model could be considered time-varying, where changes in excitability might result in different connectivity kernel parameters. A possible way to achieve this would be to employ window-based estimation, where the estimation is performed on successive windows of data (possibly overlapping windows). The window length will need to be sufficiently short such that the parameters are approximately stationary of the estimation duration, and sufficiently long to achieve a good estimation result. If this trade-off can not be successfully achieved then the level of detail in the model must be increased. A real-time implementation can also be achieved using the UKF (without the smoother\parham{???}) \parham{and an online recursive LS estimator (see \citet{Ljung1999} for example)}. }

>>>>

For models to be clinically viable, they must be patient-specific. A possible approach to achieve this would be to fit a general continuum neural field model, like the \citet{Wilson1973} (WC) or \citet{Amari1977} models, or a neural mass model like the \citet{Jansen1995} model, to patient-specific data. Fitting the neural models to individuals is a highly non-trivial task. Recently, however, this task has been approached from a number of standpoints. \dean{(Reviewer 1, additional comment d):) The first paper on patient specific modeling came from~\citet{Valdes1999}, where they fit a neural mass model of \citet{LopesDaSilva1976} and \citet{Zetterberg1978} using an extended Kalman filter. This paper demonstrated for the first time the feasibility of assimilating data with neural models of EEG.}

@@

\dean{The support of the connectivity kernel can be inferred using spatial correlation analysis. The method presented in this paper is based on a similar published method, where the kernel support was estimated using correlation analysis for a linear system described by an IDE~\citep{Scerri2009}. To demonstrate the method for the neural field model we assume the sensors are infinitesimally close (continuous observation). The spatial cross-correlation function between consecutive observations is defined as 

@@

where $\mathcal{F}^{-1}$ denotes the inverse Fourier transform and $R_{y_{t},y_{t}}(\boldsymbol{\tau})$ is the auto-correlation of the observations, $\delta_K(\cdot)$ is the Kronecker delta function, and $\delta_D(\cdot)$ is the Dirac delta function. We define the auto-correlation of the observations as

@@

This equation states that the support of the kernel can be estimated given knowledge of $\sigma_\varepsilon^2$. Furthermore, is we assume a local coupling (at $\boldsymbol\tau=0$) regime, either excitation or inhibition, than we don't even need to know $\sigma_\varepsilon^2$. The peak in the kernel at $\mathbf{r}=0$ will be bigger or smaller if we guess incorrectly and the kernel parameters can do the rest! }

@@

The sigma vectors are propagated through the system equations and weighted to form the predicted mean and covariance. The weights are calculated by 
\begin{align}
	\mathbf W_0^{(m)}&=\frac{\lambda}{ n_x+\lambda} \\
	\mathbf W_0^{(c)}&=\frac{\lambda}{ n_x+\lambda}+(1-\alpha^2+\beta) \\
	\mathbf W_i^{(m)}&=\mathbf W_i^{(c)}=\frac{1}{2( n_x+\lambda)} \quad i=1, \dots, 2n_x, 
\end{align}
where the superscripts $m$ and $c$ denote mean and covariance and $\beta$ incorporates prior knowledge of the distribution of the states, $\mathbf{x}$ \dean{(for a Gaussian disturbance $\beta$ should be set to 2)}. Since the observation equation is linear (equation~\ref{ObservationEquation}), the standard Kalman Filter update equations are used to correct the predicted states. The state estimates from forward filtering are used to form a new set of sigma points for the smoother, as described above. The cross-covariance matrix of the states, $\mathbf M$, is also required for computing the smoother gain, as described in the summary of the steps in the URTSS algorithm is given in Table~\ref{tab:UKFAlgorithm}.

Although the system is nonlinear, the parameters of the system are linear with respect to the state. This is exploited by our procedure where the parameter estimation uses a least squares (LS) method that minimizes the sum of the squared errors (of a predicted state update) with each new estimate of the state vector sequence. To create the least squares parameter estimator, we first define the $n_x \times n_{\theta}$ matrix
\begin{equation}
	\mathbf{q}(\mathbf{x}_t) = \int_\Omega \boldsymbol{\Psi}(\mathbf{r}') f(\boldsymbol{\phi}^{\top}(\mathbf{r}')\mathbf{x}_t) \,\textrm{d}\mathbf{r}'.
\end{equation}
Now given a state estimate sequence from the initialization or an iteration of the URTSS, we can write
\begin{align*}
	 \hat{\mathbf x}^b_{1} &= \mathbf{q}(\hat{\mathbf x}^b_0) \boldsymbol{\theta}+\xi\hat{\mathbf x}^b_0+\mathbf e_0 \\
	\hat{\mathbf x}^b_{2} &= \mathbf{q}(\hat{\mathbf x}^b_1) \boldsymbol{\theta}+\xi\hat{\mathbf x}^b_1+\mathbf e_1  \\
	&\vdots& \\
	\hat{\mathbf x}^b_{T}&= \mathbf{q}(\hat{\mathbf x}^b_{T-1}) \boldsymbol{\theta}+\xi\hat{\mathbf x}^b_{T-1}+\mathbf e_{T-1}. 
\end{align*}
This can be written in the compact form
\begin{equation}
	\mathbf Z=\mathbf X \mathcal W+\mathbf{e}, 
\end{equation}
where
\begin{small}
\begin{equation*}
	\mathbf Z=\left[
	\begin{array}{cccc}
		\hat{\mathbf x}^b_{1}\\
		\hat{\mathbf x}^b_{2}\\
		\vdots\\
		\hat{\mathbf x}^b_{T}
	\end{array}
	\right],\quad \mathbf X=\left[
	\begin{array}{cccc}
		\mathbf q(\hat{\mathbf x}^b_0)& \hat{\mathbf x}^b_{0}\\
		\mathbf q(\hat{\mathbf x}^b_1)& \hat{\mathbf x}^b_{1}\\
		\vdots & \vdots\\
		\mathbf q(\hat{\mathbf x}^b_{T-1})& \hat{\mathbf x}^b_{T-1}
	\end{array}
	\right] 
\end{equation*}
\end{small}
and
\begin{small}
\begin{equation*}
\quad \mathcal W=\left[
	\begin{array}{cc}
		\boldsymbol{\theta} \\
		\xi
	\end{array}
	\right],\quad \mathbf{e}=\left[
	\begin{array}{cccc}
		\mathbf e_0\\
		\mathbf e_1\\
		\vdots\\
		\mathbf e_{T-1}
	\end{array}
	\right].
\end{equation*}
\end{small}
Following this, the LS parameter estimates \citep{Ljung1985}, $ \hat{\mathcal{W}}$, are
\begin{equation}
	\hat{\mathcal{W}}=(\mathbf X^\top\mathbf X)^{-1}\mathbf X^\top\mathbf Z. 
\end{equation}

@@

\dean{The framework assumes that the parameters of the model (and brain) are stationary over the estimation period. This is a reasonable assumption if applying the framework to an evoked-potential experiment over a short time period in a controlled environment. However, it is well known that factors that influence excitability are non-stationary over longer time periods (see \citet{Ullah2010} for a nice example). Therefore, the estimation framework must be altered if it is to be applied to data over extended periods of time. A possible way to achieve this is to use a more sophisticated model that accounts for longer term time dependancies (higher temporal order). Alternatively, the parameters of the model could be considered time-varying, where changes in excitability might result in different connectivity kernel parameters. A possible way to achieve this would be to employ window-based estimation, where the estimation is performed on successive windows of data (possibly overlapping windows). The window length will need to be sufficiently short such that the parameters are approximately stationary of the estimation duration, and sufficiently long to achieve a good estimation result. If this trade-off can not be successfully achieved then the level of detail in the model must be increased. A real-time implementation can also be achieved using the UKF (without the smoother) and a real-time LS estimator (see REF for example). }
